---
title: "Case study 1: Constant growth with SUSTAIN Trout data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{constant-growth}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align='center'
)
```

## Overview

In circumstances where the number of observations available per individual is very limited, average growth rates over time may be the only plausible model to fit. In particular, if there are individuals with only two size observations, than the best that can be done is a single estimate of growth rate based on that interval. Such a model behaves as **constant growth**, which we can think of as the average rate of change across the observation period and is given by
\begin{equation}\tag{3}\label{eqn_3_const}
f(Y(t), \beta) = \beta
\end{equation}
where $\beta$ is the average growth rate. The constant growth model corresponds to linear sizes over time, and is equivalent to a linear mixed model for size, where there is an individual effect when fit to multiple individuals.

Let's simulate some data to visualise the constant growth function.  

### Load dependencies

```{r setup}
# remotes::install_github("traitecoevo/hmde")
# install.packages(c("dplyr", "ggplot2"))

library(hmde)
library(dplyr)
library(ggplot2)
```

### Visualise data

Here are some plots to demonstrate how the constant growth function relates to sizes over time for a single individual. Feel free to play around with the parameter settings (`beta`, `y_0`) and see how the plot changes.

```{r, fig.width=3, fig.height=3}
#Simulate data
beta <- 2 #Annual growth rate
y_0 <- 1 #Starting size
time <- c(0,20) 
sizes_over_time <- tibble(Y_t = 1 + beta*time, #Linear sizes over time
                          t = time)
#Plot of growth function
constant_growth_function <- ggplot() +
  xlim(y_0, max(sizes_over_time$Y_t)) +
  ylim(0, beta*2) +
  labs(x = "Y(t)", y = "f", title = "Constant growth") +
  theme_classic() +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=18,face="bold")) +
  geom_function(fun=hmde_model_des("constant_single_ind"), 
                args=list(pars = list(beta)),
                colour="green4", linewidth=1,
                xlim=c(y_0, max(sizes_over_time)))

#Sizes over time
sizes_over_time <- ggplot(data = sizes_over_time, aes(x=t, y = Y_t)) +
  geom_line(colour="green4", linewidth=1) +
  xlim(0, max(sizes_over_time$t)) +
  ylim(0, max(sizes_over_time$Y_t)*1.05) +
  labs(x = "Time", y = "Y(t)", title = "Size over time") +
  theme_classic() +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=18,face="bold"))

constant_growth_function
sizes_over_time
```

A key take-away of the function plot is the relationship to what we think of as a "reasonable growth model". We don't expect constant growth rates to be realistic, at best they represent the average rate of change over a period. More complex models may be more realistic, but in this case study we are only interested in different mechanisms of size dependence, we do not use environmental covariates for example.

## SUSTAIN trout data
Our example data for the constant model comes from @moe2020_TroutData, a publicly available dataset of mark-recapture data for _Salmo trutta_ in Norway. The time between observations is not controlled, nor is the number of observations per individual. As a result the data consists primarily of individuals with two observations of size, constituting a single observation of growth which limits the growth functions that can be fit to individuals as a single parameter model is the best that can be fit to two sizes. The constant growth function in Equation \eqref{eqn_3_const} is the most appropriate of the functions we have in `hmde`, as we can interpret the single growth interval as an estimate of the average growth rate that gets fit to $\beta$.

In order to best reflect the survey data we took a stratified sample of individuals grouped by the number of available observations. We have 25 fish with two observations, 15 with three, 10 with four, for a total sample size of 50. This data is included with `hmde`

```{r}
Trout_Size_Data
```

As initial exploration we will have a look at the distribution of observed sizes, growth behaviour, and observation intervals. First we transform the data to extract growth increment and observation interval information, then plot it.
```{r}
Trout_Size_Data_transformed <- Trout_Size_Data %>%
  group_by(ind_id) %>%
  mutate(
    delta_y_obs = y_obs - lag(y_obs),
    obs_interval = time - lag(time),
    obs_growth_rate = delta_y_obs/obs_interval
  ) %>%
  ungroup()
```

```{r}
histogram_func <- function(data, var, main, xlab, ...){
  ggplot2::ggplot(data = data, aes(x = {{var}})) + 
  geom_histogram(colour = "black", fill = "grey", ...) + 
  labs(title = main,
       xlab = xlab) +
  theme_classic()
}

histogram_func(Trout_Size_Data, y_obs, 
               "Observed size distribution", "Size (cm)", binwidth = 5)

histogram_func(Trout_Size_Data_transformed, obs_interval, 
               "Observed interval distribution", "Time (yr)", binwidth = 0.55)

histogram_func(Trout_Size_Data_transformed, delta_y_obs, 
               "Observed growth increments", "Growth increment (cm)", binwidth = 5.5)

histogram_func(Trout_Size_Data_transformed, obs_growth_rate, 
               "Observed annualised growth rate distribution", "Growth rate (cm/yr)", binwidth = 80)
```


```{r, eval=FALSE}
hist(Trout_Size_Data$y_obs, 
     main = "Observed size distribution",
     xlab = "Size (cm)")

hist(Trout_Size_Data_transformed$obs_interval, 
     main = "Observation interval distribution",
     xlab = "Time (yr)")

hist(Trout_Size_Data_transformed$delta_y_obs, 
     main = "Observed growth increments",
     xlab = "Growth increment (cm)")

hist(Trout_Size_Data_transformed$obs_growth_rate, 
     main = "Observed annualised growth rate distribution",
     xlab = "Growth rate (cm/yr)")
```
The growth histograms show that there's a number of negative growth increments, some reasonably extreme, and when combined with some short observation periods we get very extreme estimates of growth rates. We can further investigate these if needed. The constant growth model assumes non-negative growth and uses a log-normal distribution for $\beta$, which will eliminate those increments from the estimated sizes. We consider eliminating negative growth biologically reasonable as we don't expect the length of fish to decrease over time, even if their mass or width might.

Now we will actually fit the model and extract the estimates. As the provided trout data is already in the form required by the hmde_assign_data function we don't need to do any further re-naming and can pass it directly.
```{r}
trout_constant_fit <- hmde_model("constant_multi_ind") |>
  hmde_assign_data(data = Trout_Size_Data)  |>
  hmde_run(chains = 4, cores = 1, iter = 2000)

trout_constant_estimates <- hmde_extract_estimates(model = "constant_multi_ind",
                                 fit = trout_constant_fit,
                                 input_measurement_data = Trout_Size_Data)
```

At the level of sizes over time we can have a look at the distribution of estimated sizes, estimated growth increments, and annualised growth rates. The negative increments are gone without actually removing those from the data set.
```{r}
measurement_data_transformed <- trout_constant_estimates$measurement_data %>%
  group_by(ind_id) %>%
  mutate(
    delta_y_obs = y_obs - lag(y_obs),
    obs_interval = time - lag(time),
    obs_growth_rate = delta_y_obs/obs_interval,
    delta_y_est = y_hat - lag(y_hat),
    est_growth_rate = delta_y_est/obs_interval
  ) %>%
  ungroup()

hist(measurement_data_transformed$y_hat, 
     main = "Estimated size distribution",
     xlab = "Size (cm)")
hist(measurement_data_transformed$delta_y_est, 
     main = "Estimated growth increments",
     xlab = "Growth increment (cm)")
hist(measurement_data_transformed$est_growth_rate, 
     main = "Estimated annualised growth rate distribution",
     xlab = "Growth rate (cm/yr)")
```
We can also directly compare the observed sizes over time to estimated values. Quantitatively we can use statistics such as $R^2$ calculated on $(y_{ij}, \hat{Y}_{ij})$, and qualitatively we can look at plots of sizes over time. We look at 5 individuals to start with because the joint plot of sizes over time can get very messy very fast.
```{r}
#Quantitative R^2
cor(measurement_data_transformed$y_obs, measurement_data_transformed$y_hat)^2

#Plots of size over time for a sample of 5 individuals
sample_ids <- sample(1:nrow(trout_constant_estimates$individual_data), size=5)
plot_data <- measurement_data_transformed %>%
  filter(ind_id %in% sample_ids)

ggplot(data=plot_data, aes(group = ind_id)) +
  geom_point(aes(x = time, y=y_obs, colour = as.factor(ind_id)), 
             shape = 1) +
  geom_line(aes(x = time, y=y_obs, colour = as.factor(ind_id)), 
            linetype = "dashed") +
  geom_point(aes(x = time, y=y_hat, colour = as.factor(ind_id)), 
             shape = 2) +
  geom_line(aes(x = time, y=y_hat, colour = as.factor(ind_id)), 
            linetype = "solid") +
  labs(x="Time (years)", y="Size (cm)", colour="Ind. ID") +
  theme_classic()
```

At the level of individuals we are interested in the distribution of $\beta$ estimates, which will align with the estimated annualised growth rates as that's precisely what they represent. We also provide an easy way to produce a plot of the fitted growth functions in order to see how they compare to the observed sizes.
```{r}
hist(trout_constant_estimates$individual_data$ind_beta_mean,
     main = "Individual beta parameters", 
     xlab = "beta estimate")

hmde_plot_de_pieces(model = "constant_multi_ind",
                    individual_data = trout_constant_estimates$individual_data,
                    measurement_data = trout_constant_estimates$measurement_data)
```

We also get estimates of the population-level hyper-parameters that govern the distribution of $\beta$ -- $\mu$ and $\sigma$ for the log-normal distribution. These are calculated in the context of the log-transformed parameters so the easiest way to interpret $\mu$ is to back-transform it through exponentiation, but this does not so easily transfer to $\sigma$. The CIs in this output are posterior credible intervals taken as the central 95% quantiles of the posterior samples. 
```{r}
#Mean of normal distribution
trout_constant_estimates$population_data$mean[1] #Raw value
print(paste0("95% CI for mean log growth: (", 
             trout_constant_estimates$population_data$CI_lower[1], " , ",
             trout_constant_estimates$population_data$CI_upper[1], ")")) #Raw CI

exp(trout_constant_estimates$population_data$mean[1]) #In cm/yr units
print(paste0("95% CI for mean growth in cm/yr: (", 
             exp(trout_constant_estimates$population_data$CI_lower[1]), " , ",
             exp(trout_constant_estimates$population_data$CI_upper[1]), ")"))

#Standard deviation of underlying normal distribution
trout_constant_estimates$population_data$mean[2]
print(paste0("95% CI for log growth standard deviation: (", 
             trout_constant_estimates$population_data$CI_lower[2], " , ",
             trout_constant_estimates$population_data$CI_upper[2], ")")) #Raw CI
```
From the species-level data we can say that the average annual growth rate for the species is estimated to be 2.4cm/yr, with a 95% posterior CI of (1.83, 3.04). As we fit a constant growth model there's only so much we can say about the growth behaviour. 
