---
title: "Case study 2: von Bertalanffy growth with lizard size data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{von-bertalanffy}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# remotes::install_github("traitecoevo/hmde")
# install.packages(c("dplyr", "ggplot2"))

library(hmde)
library(dplyr)
library(ggplot2)
```

Our second demo introduces size-dependent growth based on the von Bertalanffy function
$$f(Y(t); S_{max}, \beta) = \beta(S_{max} - Y(t)),$$
where $S_{max}$ is the asymptotic maximum size and $\beta$ controls the growth rate. We have implemented the analytic solution
$$Y(t) = S_{max} + (Y(0) - S_{max}) \exp(-t\beta)$$
which is independent of age at the starting size $Y(0)$ and instead uses the first size as the initial condition. The key behaviour of the von Bertalanffy model is a high growth rate at small sizes that declines linearly as the size approaches $S_{max}$. This manifests as growth slowing as a creature matures with a hard finite limit on the eventual size. We restrict $\beta$ and $S_{max}$ to be positive, and $S_{max}$ to be larger than the observed sizes. As a result the growth rate is non-negative. 

In the following code we plot an example of the growth function and the solution to get a feel for the behaviour.
```{r}
#Analytic solution in function form
solution <- function(t, pars = list(y_0, beta, S_max)){
  return(
    pars$S_max + (y_0 - pars$S_max)*exp(-t * pars$beta)
  )
}

#Parameters
beta <- 0.35 #Growth rate
y_0 <- 1 #Starting size
S_max <- 20 #Asymptotic max size
time <- c(0,30) 
pars_list <- list(y_0 = y_0,
                  beta = beta,
                  S_max = S_max)
y_final <- solution(time[2], pars_list)

#Plot of growth function
ggplot() +
  xlim(y_0, y_final) +
  ylim(0, beta*(S_max-y_0)*1.1) +
  labs(x = "Y(t)", y = "f", title = "von Berralanffy growth") +
  theme_classic() +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=18,face="bold")) +
  geom_function(fun=hmde_model_des("vb_single_ind"), 
                args=list(pars = list(S_max, beta)),
                colour="green4", linewidth=1,
                xlim=c(y_0, y_final))

#Size over time
ggplot() +
  geom_function(fun=solution, 
                args=list(pars = pars_list),
                colour="green4", linewidth=1,
                xlim=c(time)) +
  xlim(time) +
  ylim(0, y_final*1.05) +
  labs(x = "Time", y = "Y(t)", title = "von Bertalanffy growth") +
  theme_classic() +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=18,face="bold"))

```

The von Bertalanffy model is commonly used in fishery management \citep{flinn2021trends}, but has also been used in reptile studies such as \cite{edmonds2021growing, zhao2020age}.

## Lizard size data
Our data is sourced from \cite{Kar2023_LizardMass} which measured mass and snout-vent-length (SVL) of delicate skinks -- \textit{Lampropholis delicata} -- under experimental conditions to examine the effect of temperature on development. We are going to use the SVL metric for size.

We took a simple random sample without replacement of 50 individuals with at least 5 observations each. The von Bertalanffy model can be fit to shorter observation lengths, but fewer than 3 observations is not advised as there are two growth parameters per individual. 

## Implementation
The workflow for the second example is the same as the first, with the change in model name and data object.
```{r}
lizard_vb_fit <- hmde_model("vb_multi_ind") |>
  hmde_assign_data(data = Lizard_Size_Data)  |>
  hmde_run(chains = 4, cores = 1, iter = 2000)

lizard_vb_estimates <- hmde_extract_estimates(model = "vb_multi_ind",
                                 fit = lizard_vb_fit,
                                 input_measurement_data = Lizard_Size_Data)
```

As before, we can compare the observed sizes over time to those predicted by the model.
```{r}
measurement_data_transformed <- lizard_vb_estimates$measurement_data %>%
  group_by(ind_id) %>%
  mutate(
    delta_y_obs = y_obs - lag(y_obs),
    obs_interval = time - lag(time),
    obs_growth_rate = delta_y_obs/obs_interval,
    delta_y_est = y_hat - lag(y_hat),
    est_growth_rate = delta_y_est/obs_interval
  ) %>%
  ungroup()

#Distributions of estimated growth and size
hist(measurement_data_transformed$y_hat, 
     main = "Estimated size distribution",
     xlab = "Size (cm)")
hist(measurement_data_transformed$delta_y_est, 
     main = "Estimated growth increments",
     xlab = "Growth increment (cm)")
hist(measurement_data_transformed$est_growth_rate, 
     main = "Estimated annualised growth rate distribution",
     xlab = "Growth rate (cm/yr)")

#Quantitative R^2
cor(measurement_data_transformed$y_obs, measurement_data_transformed$y_hat)^2

#Plots of size over time for a sample of 5 individuals
sample_ids <- sample(1:nrow(lizard_vb_estimates$individual_data), size=5)
plot_data <- measurement_data_transformed %>%
  filter(ind_id %in% sample_ids)

ggplot(data=plot_data, aes(group = ind_id)) +
  geom_point(aes(x = time, y=y_obs, colour = as.factor(ind_id)), 
             shape = 1) +
  geom_line(aes(x = time, y=y_obs, colour = as.factor(ind_id)), 
            linetype = "dashed") +
  geom_point(aes(x = time, y=y_hat, colour = as.factor(ind_id)), 
             shape = 2) +
  geom_line(aes(x = time, y=y_hat, colour = as.factor(ind_id)), 
            linetype = "solid") +
  labs(x="Time (days)", y="Size (mm)", colour="Ind. ID") +
  theme_classic()
```

We have two parameters at the individual level and are interested in both their separate distributions, and if we see evidence of a relationship between them. We can also use the individual parameter estimates and estimated sizes to plot the growth function pieces.
```{r}
#1-dimensional parameter distributions
hist(lizard_vb_estimates$individual_data$ind_max_size_mean,
     main = "Individual max size parameters", 
     xlab = "Smax estimate")

hist(lizard_vb_estimates$individual_data$ind_growth_rate_mean,
     main = "Individual growth rate parameters", 
     xlab = "beta estimate")

#2-dimensional parameter distribution
ggplot(data = lizard_vb_estimates$individual_data, 
       aes(x = ind_max_size_mean, y = ind_growth_rate_mean)) +
  geom_point(shape = 16, size = 1, colour = "green4") +
  xlab("Individual max sizes (mm)") +
  ylab("Individual growth rate parameters") +
  theme_classic()

#Correlation of parameters
cor(lizard_vb_estimates$individual_data$ind_max_size_mean,
    lizard_vb_estimates$individual_data$ind_growth_rate_mean)

#Plot function pieces over estimated sizes.
hmde_plot_de_pieces(model = "vb_multi_ind",
                    individual_data = lizard_vb_estimates$individual_data,
                    measurement_data = lizard_vb_estimates$measurement_data)
```

At the hyper-parameter level for the whole population we have centre and spread parameters for the log-normal distributions of $S_{max}$ and $\beta$. As before, we can look at these as species-level features.
```{r}
#Max size
lizard_vb_estimates$population_data$mean[1] #Raw value
print(paste0("95% CI for mean log max size: (", 
             lizard_vb_estimates$population_data$CI_lower[1], " , ",
             lizard_vb_estimates$population_data$CI_upper[1], ")")) #Raw CI

exp(lizard_vb_estimates$population_data$mean[1]) #In mm units
print(paste0("95% CI for mean max size in mm: (", 
             exp(lizard_vb_estimates$population_data$CI_lower[1]), " , ",
             exp(lizard_vb_estimates$population_data$CI_upper[1]), ")"))

#Standard deviation of underlying normal distribution
lizard_vb_estimates$population_data$mean[2]
print(paste0("95% CI for log max size standard deviation: (", 
             lizard_vb_estimates$population_data$CI_lower[2], " , ",
             lizard_vb_estimates$population_data$CI_upper[2], ")")) #Raw CI

#Beta
lizard_vb_estimates$population_data$mean[3] #Raw value
print(paste0("95% CI for mean log growth par: (", 
             lizard_vb_estimates$population_data$CI_lower[3], " , ",
             lizard_vb_estimates$population_data$CI_upper[3], ")")) #Raw CI

exp(lizard_vb_estimates$population_data$mean[3]) #In cm/yr units
print(paste0("95% CI for mean growth par mm/yr: (", 
             exp(lizard_vb_estimates$population_data$CI_lower[3]), " , ",
             exp(lizard_vb_estimates$population_data$CI_upper[3]), ")"))

#Standard deviation of underlying normal distribution
lizard_vb_estimates$population_data$mean[4]
print(paste0("95% CI for log growth par standard deviation: (", 
             lizard_vb_estimates$population_data$CI_lower[4], " , ",
             lizard_vb_estimates$population_data$CI_upper[4], ")")) #Raw CI

```
